---
title: "Data Manipulation with dplyr"
author: "Raee Bhagat"
date: "2022-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Manipulation with `dplyr` Practice

This homework is centered around manipulating a data set using functions from the `dplyr` package. You should be able to accomplish everything in the homework using some form of the following functions:

function | description | example |
|:---|:------|:---------|
|`filter()` | select rows based on some criteria | data %>% filter(age > 20 & sex == "f")|
|`arrange()` | sort rows | data %>% arrange(age, height) |
|`select()` | select columns | data %>% select(sex, weight) OR 
| | | data %>% select(-height) |
|`rename()` | rename columns | data %>% rename(newName = oldName) |
|`mutate()` | create new columns | data %>% mutate(old = age > 40) |
|`case_when()`| recode values | data %>% mutate(ageCat = case_when(age < 20 ~ "young", age < 40 ~ "middle", T~"old")) |
|`group_by()` | create groups for calculations | data %>% group_by(sex) |
|`summarize()` | calculate statistics on *groups* | data %>% group_by(sex) %>% summarize(meanW = mean(weight)) |
|`left_join()` | merge data sets on some ID | data %>% left_join(data2, by = "id")

## Data for the exercises

We will be using data sets for the following exercises.  The first of these comes from the `speff2trial` package. To get the data do the following:

```{r}
library(dplyr)
library(tidyverse)
library(speff2trial)
```

If you wish to see a description of the data set use `?ACTG175`. The second set of data are in the Github folder and should be downloaded when you pull the latest version; the file with the data is called `"patient_demo.csv"`. To load the data, make sure your current working directory (use `getwd()` and `setwd()` to check or change, respectively) contains the CSV file then run:

```{r}
trial <- ACTG175
demo_data <- read.csv("patient_demo.csv")
```

### `rename()`

1. Rename the columns `wtkg` to `weightkg` and `age` to `age_years`.
```{r}
trial %>% rename(weightkg = wtkg, age_years = age) -> t1
```

### `mutate()`

2. Create a column called `agem` that displays a patient's age in months rather than years.

3. The column `karnof` shows each patient's Karnofsky score on a scale of 0 to 100. Create a new column called `karnof_b` that has a scale of 0 to 1 instead.

4. Create a column called `sparrow` that has the formula of Karnofsky score/(age + weight).

```{r}
t1 %>% mutate(agem = age_years*12) -> t2

t1 %>% mutate(karnof_b = karnof/100) -> t3

t1 %>% mutate(sparrow = karnof/(age_years + weightkg)) -> t4
```

### `arrange()`

5. Sort the `trial` data in ascending values of age, weight, and sparrow, respectively. Verify that the ordering worked.

6. Using the `desc()` function, sort the trial data in order of *descending* weight.

```{r}
t4 %>% arrange(age_years, weightkg, sparrow) -> t5
t5 %>% slice_head(n=10)

t5 %>% arrange(desc(weightkg)) -> t6
t6 %>% slice_head(n=10)
```

### `case_when()`

7. In one *single* call, create two columns in the data: the first creates a column named `gender_char` that shows gender as either "male" or "female" and the second column creates a discrete variable `over50` that is 0 when the patient is younger than 50 and 1 otherwise.

```{r}
t4 %>% mutate(gender_char = case_when(gender == "0" ~ "female", gender == "1" ~ "male"),
               dv = case_when(age_years < "50" ~ 0, age_years >= "50" ~ 1)) -> t7
```

### `filter()`

8. Create a new data set `trial_men` that only had data from males.

9. Create another data set `trial_sub` that only has individuals that are: female, haven't used intravenous drugs, and are over 40 years old. 

```{r}
t7 %>% filter(gender_char == "male") -> trial_men

t7 %>% filter(gender_char == "female") %>%
  filter(age_years > 40) %>%
  filter(drugs == "0")-> trial_sub
```

### `select()`

10. Drop the `karnof` and `weight` columns from the `trial_men` data and save the results.

11. Keep only the `pidnum`,  `treat`, and `cd820` columns from `trial_sub` and save the results.

```{r}
trial_men %>% select(-karnof, -weightkg) -> t10

trial_sub %>% select(pidnum, treat, cd820) -> t11
```

### `left_join()`

12. Merge the `trial` and `demo_data` data sets. Save the resulting data as `trial_full`. 

13. Using the `trial_full` data, calculate the mean number of days of exercise reported by patients.

```{r}
trial %>% left_join(demo_data) -> trial_full #joining, by = "pidnum"

trial_full %>% summarise(mean_exercise = mean(exercise, na.rm = T)) -> mean_exercise
```

14. Group the data by `arms`. Then, for each arm, calculate the mean participant age.

15. Group the data by `arms`. For each arm, calculate the mean participant age and the median Karnofsky score.

16. Group the data by `gender`. Then, separately for male and female patients, calculate the percent who have a history of intravenous drug use. (To calculate a percent of a binary variable with 0s and 1s, just calculate the mean)

17. Group the data by `gender`. Then calculate the percent of male and female patients who have a history of intravenous drug use (drugs), and the mean number of days until a major negative event days

18. Separately for all combinations of `gender` and `race`, calculate the mean age and mean CD4 T cell count at baseline (`cd40`). (Hint: group by gender and race.)

```{r}
trial_full %>% group_by(arms) %>% summarize(mean_age = mean(age))-> t14

trial_full %>% group_by(arms) %>% summarize(mean_age = mean(age),
                                            med_karnofsky = median(karnof))-> t15

trial_full %>% group_by(gender) %>% summarize(per_In = mean(drugs == 1)) -> t16 
 
trial_full %>% group_by(gender) %>% summarize(per_In = mean(drugs == 1),
                                               mean_major_neg = mean(days)) -> t17

trial_full %>% group_by(gender, race) %>% summarize(mean_age = mean(age),
                                                   mean_cd4 = mean(cd40)) -> t18
```
#Exercises

19. Now let’s check the major differences between the treatment arms. For each arm, calculate the following:
   - Mean days until a a major negative event (days)
   - Mean CD4 T cell count at baseline.
   - Mean CD4 T cell count at 20 weeks.
   - Mean CD4 T cell count at 96 weeks.
   - Mean change in CD4 T cell count between baseline and 96 weeks
   - Number of patients (Hint: use `N = n()`)

```{r}
trial_full %>% group_by(arms) %>% summarize(mean_major_neg = mean(days),
                                            mean_cd4_baseline = mean(cd40),
                                            mean_cd4_20 = mean(cd420),
                                            mean_cd4_96 = mean(cd496, na.rm = T),
                                            mean_change = mean_cd4_baseline - mean_cd4_96,
                                            N = n()) -> t19
```

20. Repeat the previous analysis, but before you do the grouping and summary statistics, create a new variable called `arms_char` that shows the values of `arms` as characters that reflect what the values actually represent (hint: use `mutate()` and `case_when()`). For example, looking at the help file `?ACTG175`, I can see that the treatment arm of 0 is “zidovudine”. I might call this arm `"Z"`. Do this **all in the same chunk of code**.

```{r}
t19 %>% mutate(arms_char = case_when(arms == "0" ~ "zidovudine",
                                            arms == "1" ~ "zidovudine and didanosine",
                                            arms == "2" ~ "zidovudine and zalcitabine",
                                            arms == "3" ~ "didanosine")) -> t20
```

21. Repeat the previous analysis, but only include patients with a Karnofsky score equal to 100, and who did not use zidovudine in the 30 days prior to the treatment initiation (`z30`).

```{r}
trial_full %>% filter(karnof == 100, z30 == "0") %>% 
  group_by(arms) %>%
  summarize(mean_major_neg = mean(days),
            mean_cd4_baseline = mean(cd40),
            mean_cd4_20 = mean(cd420),
            mean_cd4_96 = mean(cd496, na.rm = T),
            mean_change = mean_cd4_baseline - mean_cd4_96,
            N = n()) -> t21
```

22. In **one block of code**, complete the following tasks
   - Change the name of the column `weightkg` to `weight`
   - Create a new column called `drugs_char` that uses strings instead of numbers to indicate drug use
   - Filter the data to only include male patients with id numbers greater than 10100
   - Group the data by `drugs_char` and `arms`
   - For each group calculate the mean age, percentage of patients that are male, and mean number of days until a major negative event
   
```{r}
trial_full %>% rename(weight = wtkg) %>%
  mutate(drugs_char =case_when(drugs == "0" ~ "No Intravenous drug",
                               drugs == "1" ~ "Use of Intravenous drug")) %>%
  filter(gender == "1", pidnum > 10100) %>%
  group_by(drugs_char, arms) %>%
  summarize(mean_age = mean(age),
            per_patients_male = mean(gender)*100,
            mean_major_neg = mean(days))-> t22
```





````````````````````````````````````````````````````````````````````````````````````
